Empirical Analysis of the Role of Energy in Economic Growth

### 2013-01-21 EconGrowth.Rnw
Now sending column name string and text to be used in tables into the C-D with energy function.

<<cobbDouglasEnergy function>>=
###############################
# This function fits the Cobb Douglas production function including energy to historical data
# The functional form is
#
# iGDP = exp(lambda * itime) * iCapStk^alpha * iLabor^beta * iEnergy^gamma
# 
# iYear: time indexed to 0.0 at the first year [years since beginning of data set]
# iGDP = time series GDP data indexed to 1.0 at the first year [-]
# iCapStk = time series capital stock data indexed to 1.0 at the first year [-]
# iLabor = time series labor data indexed to 1.0 at the first year [-]
# iEnergy = time series energy data indexed to 1.0 at the first year [-]. 
#      This may be any of heat (iQ), exergy (iX), or useful work (iU).
# 
# returns a data object containing three rows and 5 columns.
#   col 1: lambda
#   col 2: alpha
#   col 3: beta
#   col 4: gamma
#   row 1: -95% CI for each parameter
#   row 2: estimate for each parameter
#   row 3: +95% CI for each parameter
#
# dataTable is the table to be used for this curve fit
# iEnergy is the name of the column to be used for energy.
#   This may be any of heat (iQ), exergy (iX), or useful work (iU).
##
cobbDouglasEnergy <- function(dataTable, iEnergy){
  # Reparameterize to ensure that we meet the constraints:
  # * alpha + beta + gamma = 1.0.
  # * alpha, beta, and gamma are all between 0.0 and 1.0.
  # To do this, we reparameterize as
  # * 0 < a < 1
  # * 0 < b < 1
  # * alpha = min(a, b)
  # * beta = b - a
  # * gamma = 1 - max(a, b)
  energyName = names(iEnergy) #grabs the name of the energy column
  lambdaGuess <- 0.0 # guessing lambda = 0 means there is no technological progress.
  alphaGuess <- 0.3 # a typical value for alpha
  betaGuess <- 0.7 # a typical value for beta
  modelCDe <- nls(iGDP ~ exp(lambda*iYear) * 
                    iCapStk^min(a,b) * iLabor^abs(b-a) * 
                    iEnergy^(1.0 - max(a,b)), 
                  algorithm = "port",
                  start = list(lambda=lambdaGuess, a=alphaGuess, b=alphaGuess+betaGuess),
                  lower = list(lambda=-Inf, a=0, b=0),
                  upper = list(lambda=Inf, a=1, b=1),
                  data = dataTable)
  aicCDe <- AIC(modelCDe, k=2) # Checks validity of the model. AIC stands for Akaike's Information Criterion
  print(aicCDe)

@@

  # Combine all estimates and their confidence intervals into data frames with intelligent row names
  estCDe <- data.frame(lambda = lambda, alpha = alpha, beta = beta, gamma = gamma)
  print(estCDe);
  #row.names(estCDe) <- "Cobb-Douglas with e: $y = e^{\\lambda t}k^{\\alpha}l^{\\beta}e^{\\gamma}$"
  row.names(estCDe) <- paste("Cobb-Douglas with ", energyName, sep="")
  # The [1] subscripts pick off the lower confidence interval
  lowerCDe <- data.frame(lambda = ciCDe["lambda","2.5%"], alpha = ciCDe["a", "2.5%"], beta = betaCICDe[1], gamma = gammaCICDe[1])
  row.names(lowerCDe) <- "- 95% CI"
  # The [2] subscripts pick off the lower confidence interval
  upperCDe <- data.frame(lambda = ciCDe["lambda","97.5%"], alpha = ciCDe["a", "97.5%"], beta = betaCICDe[2], gamma = gammaCICDe[2])
  row.names(upperCDe) <- "+ 95% CI"

@@

<<cdq>>=
dataCDe <- cobbDouglasEnergy(dataTable, dataTable["iQ"])
tableCDe <- xtable(dataCDe, caption="U.S. 1980-2011.", digit = c(4, 4, 2, 2, 3))
#estCDe <- dataCDe[2]
#dataAll <- rbind(estCD, estCDe); dataAll
#tableAll <- xtable(dataAll, caption="U.S. 1980-2011, All Models.", digit = c(4, 4, 2, 2, 3))
@

@@

print(tableCDe, caption.placement="top")
# According to http://cran.r-project.org/web/packages/xtable/vignettes/xtableGallery.pdf, Section 3.1, I should 
# be able to use the "sanitize.text.function" parameter to allow markup in column headers. But this next
# line is not working at the present time. --MKH, 18 Jan 2012.
# print(tableCDe, sanitize.text.function = function(x){x})

@@

<<cdx>>=
dataCDx <- cobbDouglasEnergy(dataTable, dataTable["iX"])
tableCDx <- xtable(dataCDx, caption="U.S. 1980-2011.", digit = c(4, 4, 2, 2, 3))
@

@@

<<cdu>>=
dataCDu <- cobbDouglasEnergy(dataTable, dataTable["iU"])
tableCDu <- xtable(dataCDu, caption="U.S. 1980-2011.", digit = c(4, 4, 2, 2, 3))
@

@@

An issue arrises in this example because contraining $\alpha$ and $\beta$ is not sufficient to 
guarantee that $\gamma$ is properly constrained.

<<cduold>>=
# Establish guess values for alpha, beta, and lambda.
lambdaGuess <- 0.0 # guessing lambda = 0 means there is no technological progress.
alphaGuess <- 0.2 # a typical value for alpha
betaGuess <- 0.6 # a typical value for beta
gammaGuess <- 0.01 # a nice low value

# Runs a non-linear least squares fit to the data with constraints
modelCDU <- nls(iGDP ~ exp(lambda*iYear) * iCapStk^alpha * iLabor^beta * iU^(1.0 - alpha - beta), 
                algorithm="port",
                start = list(lambda=lambdaGuess, alpha=alphaGuess, beta=betaGuess),
                lower = list(lambda=-Inf, alpha=0, beta=0),
                upper = list(lambda=Inf, alpha=1, beta=1),
                data=dataTable)

# Gives the nls summary table
summary(modelCDU)
confint(modelCDU, level = 0.95)
@

<<>>=
# Checks validity of the model. AIC stands for Akaike's Information Criterion
AIC(modelCDU, k=2)
@

The problem here is that $\hat \gamma < 0$.
<<>>=
# Calculate gamma and report it.
alpha <- coef(modelCDU)["alpha"] 
beta <- coef(modelCDU)["beta"] 
gamma <- as.numeric(1.0 - alpha - beta)
c(coef(modelCDU), gamma=gamma)
@

Our $\hat\gamma$ is not much below $0$.  Let's compute the standard error.
<<>>=
require(car)  # the methods in alr3 have been deprecated.  Use car instead.
gamma.est <- deltaMethod(modelCDU, "1-alpha-beta"); gamma.est
# crude CI (using 2 as a rough estimate for critical value)
with(gamma.est, Estimate + c(-1,1) * 2 * SE)
@
So no real cause for concern: our data don't convince us that the real $\gamma$ is different from $0$.

\subsection{Forcing $\gamma\ge0$}

<<>>=
modelCDUforced <- nls(iGDP ~ exp(lambda*iYear) * 
                        iCapStk^min(a,b) * iLabor^abs(b-a) * 
                        iU^(1.0 - max(a,b)), 
                      algorithm="port",
                      start = list(lambda=lambdaGuess, a=alphaGuess, b=alphaGuess + betaGuess),
                      lower = list(lambda=-Inf, a=0, b=0),
                      upper = list(lambda=Inf, a=1, b=1),
                      data=dataTable)

coef(summary(modelCDUforced))
with( as.data.frame(t(coef(modelCDUforced))), c(alpha=min(a,b), beta=abs(b-a), gamma=1-max(a,b)))
@
But the naive delta method to calculate significance information fails because R doesn't know how to calculate the derivatives for the minimum and maximum functions.  So we need to be clever, using the fact that we know now that $a < b$:
<<>>=
# alpha = a
deltaMethod( modelCDUforced, "a")
# beta = b - a
deltaMethod( modelCDUforced, "b-a")
# gamma = 1-b
deltaMethod( modelCDUforced, "1-b")
@

So this seems to give us what we want for this case:  We have parameter estimates and standard errors subject to all of our constraints.

It may be that we can avoid using min and max and just
use $a$, $b-a$ and $1-b$.  If that works, then this generalizes fairly easily to any number of parameters that must be bounded by 0 and 1 and sum to 1.  (Else we have to ``sort'' the dummy parameters first, which is OK but makes the coding a bit uglier.)

>>>>

<<cobbDouglasEnergy function>>=
###############################
# This function fits the Cobb Douglas production function including energy to historical data
# The functional form is
#
# iGDP = exp(lambda * itime) * iCapStk^alpha * iLabor^beta * iEnergy^gamma
# 
# iYear: time indexed to 0.0 at the first year [years since beginning of data set]
# iGDP = time series GDP data indexed to 1.0 at the first year [-]
# iCapStk = time series capital stock data indexed to 1.0 at the first year [-]
# iLabor = time series labor data indexed to 1.0 at the first year [-]
# iEnergy = time series energy data indexed to 1.0 at the first year [-]. 
#      This may be any of heat (iQ), exergy (iX), or useful work (iU).
# 
# returns a data object containing three rows and 5 columns.
#   col 1: lambda
#   col 2: alpha
#   col 3: beta
#   col 4: gamma
#   row 1: -95% CI for each parameter
#   row 2: estimate for each parameter
#   row 3: +95% CI for each parameter
#
# dataTable is the table to be used for this curve fit
# iEnergy is the name of the column to be used for energy.
#   This may be any of heat (iQ), exergy (iX), or useful work (iU).
##
cobbDouglasEnergy <- function(dataTable, energyColumnName, energyString){
  # Reparameterize to ensure that we meet the constraints:
  # * alpha + beta + gamma = 1.0.
  # * alpha, beta, and gamma are all between 0.0 and 1.0.
  # To do this, we reparameterize as
  # * 0 < a < 1
  # * 0 < b < 1
  # * alpha = min(a, b)
  # * beta = b - a
  # * gamma = 1 - max(a, b)
  iEnergy <- dataTable[energyColumnName] #grabs the energy column
  lambdaGuess <- 0.0 # guessing lambda = 0 means there is no technological progress.
  alphaGuess <- 0.3 # a typical value for alpha
  betaGuess <- 0.7 # a typical value for beta
  modelCDe <- nls(iGDP ~ exp(lambda*iYear) * 
                    iCapStk^min(a,b) * iLabor^abs(b-a) * 
                    iEnergy^(1.0 - max(a,b)), 
                  algorithm = "port",
                  start = list(lambda=lambdaGuess, a=alphaGuess, b=alphaGuess+betaGuess),
                  lower = list(lambda=-Inf, a=0, b=0),
                  upper = list(lambda=Inf, a=1, b=1),
                  data = dataTable)
  aicCDe <- AIC(modelCDe, k=2) # Checks validity of the model. AIC stands for Akaike's Information Criterion
  print(aicCDe)

@@

  # Combine all estimates and their confidence intervals into data frames with intelligent row names
  estCDe <- data.frame(lambda = lambda, alpha = alpha, beta = beta, gamma = gamma)
  print(estCDe);
  #row.names(estCDe) <- "Cobb-Douglas with e: $y = e^{\\lambda t}k^{\\alpha}l^{\\beta}e^{\\gamma}$"
  row.names(estCDe) <- paste("C-D with ", energyString, sep="")
  # The [1] subscripts pick off the lower confidence interval
  lowerCDe <- data.frame(lambda = ciCDe["lambda","2.5%"], alpha = ciCDe["a", "2.5%"], beta = betaCICDe[1], gamma = gammaCICDe[1])
  row.names(lowerCDe) <- "- 95% CI"
  # The [2] subscripts pick off the lower confidence interval
  upperCDe <- data.frame(lambda = ciCDe["lambda","97.5%"], alpha = ciCDe["a", "97.5%"], beta = betaCICDe[2], gamma = gammaCICDe[2])
  row.names(upperCDe) <- "+ 95% CI"

@@

<<cdq>>=
dataCDq <- cobbDouglasEnergy(dataTable, "iQ", "q")
# TODO Develop a better caption for the table based on the data automatically.
tableCDq <- xtable(dataCDq, caption="U.S., 1980-2011.", digit = c(4, 4, 2, 2, 3))
#estCDe <- dataCDe[2]
#dataAll <- rbind(estCD, estCDe); dataAll
#tableAll <- xtable(dataAll, caption="U.S. 1980-2011, All Models.", digit = c(4, 4, 2, 2, 3))
@

@@

print(tableCDq, caption.placement="top")
# According to http://cran.r-project.org/web/packages/xtable/vignettes/xtableGallery.pdf, Section 3.1, I should 
# be able to use the "sanitize.text.function" parameter to allow markup in column headers. But this next
# line is not working at the present time. --MKH, 18 Jan 2012.
# print(tableCDe, sanitize.text.function = function(x){x})

@@

<<cdx>>=
dataCDx <- cobbDouglasEnergy(dataTable, "iX", "x")
tableCDx <- xtable(dataCDx, caption="U.S., 1980-2011.", digit = c(4, 4, 2, 2, 3))
@

@@

<<cdu>>=
dataCDu <- cobbDouglasEnergy(dataTable, "iU", "U")
tableCDu <- xtable(dataCDu, caption="U.S., 1980-2011.", digit = c(4, 4, 2, 2, 3))
@

@@
